### 2021-06-10 memo.tex
refinements in counting zero modes

The monopoles associated to the breaking of $G$ to its Cartan were analyzed in many places, see e.g.~\cite{Gauntlett:1996cw}.\footnote{%
There were many competing papers from the same period concerning the application to Seiberg-Witten;
the analysis of the monopoles itself is much older and I should find proper original references for it too.
}
The upshot is that you can embed the standard spherically-symmetric 't Hooft-Polyakov monopole using any simple root into $G$ and have a monopole solution without additional bosonic moduli.

So we consider a monopole with the charge $(1,1,0,\ldots,0)$.
This uses $su(2)_1$ within $so(4)\simeq su(2)_1\times su(2)_2$ within $so(2n_c)\subset so(4)\times so(4n_c-4)$.

Since the fermions are in the vector representation of $so(2n_c)$,
the charged part under $su(2)_1$ is of charge 1 and in the doublet of $su(2)_2$.
Therefore it gives rise to fermionic zero modes transforming in \begin{equation}
V_{2} \otimes R_{2n_f} 
\end{equation} where $V_2$ is the doublet of $su(2)_2$, 
$R_{2n_f}$ is the fundamental of $usp(2n_f)$,
and we need to impose the reality condition using the pseudoreality of both factors,
so that there are $4n_f$ Majorana fermion in total.

I haven't analyzed the general case; let's take $n_f=1$ and $n_f=2$ as examples.
When $n_f=1$, there are $4$ Majorana fermions.
Quantizing them, we find the monopoles in \begin{equation}
V_2 \otimes \mathbf{1} \oplus \mathbf{1}\otimes R_2.
\end{equation}
%I think we should take the spacetime scalar as indicative, (which needs to be justified)
It has the `vector' charge under $usp(2)$ flavor symmetry or is a doublet under $su(2)_2$.
In either case, the $-1$ in $SO(2n_c)\times USp(2n_f)$ acts nontrivially,
and the 1-form symmetry is extended.

When $n_f=2$, there are $8$ Majorana fermions, 
and  $su(2)_2$ and flavor $usp(4)$ are embedded into $so(8)$ via \begin{equation}
su(2)\times usp(4) \simeq so(3)\times so(5) \subset so(8).
\end{equation}
Therefore, the monopoles are easily seen to be in \begin{equation}
V_3\otimes \mathbf{1} \oplus V_2\otimes R_2 \oplus V_1\otimes R_5.
\end{equation}
Again,
it is in the `adjoint' charge under the $usp(4)$ flavor symmetry.
Therefore the 1-form symmetry is not extended.

>>>>

The monopoles associated to the breaking of $G$ to its Cartan were analyzed in many places, 
e.g.~in \cite{Weinberg:1979zt}.
There, the following was shown.
Let $\phi$ be the scalar vev in the real Cartan subalgebra $\mathfrak{h}\subset \mathfrak{g}$.
This determines the simple roots $\alpha$.
Then you can embed the standard spherically-symmetric 't Hooft-Polyakov monopole and have a monopole solution without additional bosonic moduli.

Let us say we chose the standard $\phi$ such that the simple roots are $(1,-1,\ldots,0)$,
$(0,1,-1,\ldots,0)$, \ldots, $(0,\ldots,1,-1)$ and $(0,\ldots,1,+1)$.
They generate the sublattice $\Lambda$ of $\bZ^n$.
The group of the 't Hooft line operators up to the screening by the dynamical monopoles is then \begin{equation}
\bZ^n / \Lambda = \bZ_2.
\end{equation}
What we want to do is take into account the center charge $\bZ_2$ of $USp(2n_f)$. 
For this, we consider $\bZ^n\times \bZ_2$,
and divide it by the subgroup generated by the charges of the monopoles associated to the simple roots,
which we denote by $(1,-1,\ldots, 0; q_1)$, $(0,1,-1,\ldots, 0; q_2)$, \ldots,
$(0,\ldots, 1,-1; q_{n-1})$ and $(0,\ldots,1,+1; q_n)$.

We do not have to determine $q_1$ to $q_{n-1}$. We simply use them to rewrite any charge vector \begin{equation}
(m_1,\ldots, m_{n-2}, m_{n-1},m_n; q)
\end{equation} 
into \begin{equation}
(0,\ldots,0, m, m' ; q')
\end{equation}
We then have to determine $q_{n-1}$ and $q_n$.

This reduces the study to the case of $so(4)\simeq su(2)_1 \times su(2)_2$.
The monopoles associated to the simple roots are just 't Hooft-Polyakov monopoles associated to the two factors of $su(2)$'s.
The vev of the adjoint scalar in this basis can be written as $(a_1,a_2)$, which we assume $a_1>a_2>0$.
The fermion is in the vector representation of $so(4)$.
Under the monopole in $su(2)_1$, it is a doublet coupled to an adjoint vev of size $a_1$ with bare mass $a_2$,
and similarly for the monopole in $su(2)_2$.

Now, the explicit analysis in \cite[Sec.~IV]{Callias:1977kg} concerning the number of zero modes in the 't Hooft-Polyakov monopole says that 
a doublet coupled to an adjoint vev of size $a$ with bare mass $\mu$ has
 a zero mode if $|a|>|m|$ and 
 it has no zero mode if $|a|<|m|$.

With our assumption $a_1>a_2>0$, this means that the monopole in $su(2)_1$ has a zero mode while the one in $su(2)_2$ doesn't have any.
In our original basis, this means that the monopole with $(0,\ldots,1,-1;q_{n-1})$ does not have any zero mode and $q_{n-1}=0$,
while the one with $(0,\ldots,1,+1;q_n)$  (which is in $su(2)_1$) has two zero modes per flavor.

The one-form symmetry group is obtained by dividing $\bZ^2\times \bZ_2$ by the subgroup generated by $(1,-1;0)$ and $(1,+1;q_n)$.
This is $\bZ_2\times \bZ_2$ or $\bZ_4$ depending on whether $q_n$ is $0$ or $1$.

Let us determine $q_n$.
We saw two zero modes per flavor; 
this means that there are fermionic zero modes transforming in \begin{equation}
V_{2} \otimes R_{2n_f} 
\end{equation} where $V_2$ is the doublet of $su(2)_2$ (which is actually broken to $u(1)$ but keeping $su(2)_2$ representation is useful in organizing the answer), 
$R_{2n_f}$ is the fundamental of $usp(2n_f)$,
and we need to impose the reality condition using the pseudoreality of both factors,
so that there are $4n_f$ Majorana fermion in total.

I haven't analyzed the general case; let's take $n_f=1$ and $n_f=2$ as examples.
When $n_f=1$, there are $4$ Majorana fermions.
Quantizing them, we find the monopoles in \begin{equation}
V_2 \otimes \mathbf{1} \oplus \mathbf{1}\otimes R_2.
\end{equation}
%I think we should take the spacetime scalar as indicative, (which needs to be justified)
It has the `vector' charge under $usp(2)$ flavor symmetry or is a doublet under $su(2)_2$.
In either case, the $-1$ in $SO(2n_c)\times USp(2n_f)$ acts nontrivially, so $q_n=1$.
and the 1-form symmetry is extended.

When $n_f=2$, there are $8$ Majorana fermions, 
and  $su(2)_2$ and flavor $usp(4)$ are embedded into $so(8)$ via \begin{equation}
su(2)\times usp(4) \simeq so(3)\times so(5) \subset so(8).
\end{equation}
Therefore, the monopoles are easily seen to be in \begin{equation}
V_3\otimes \mathbf{1} \oplus V_2\otimes R_2 \oplus V_1\otimes R_5.
\end{equation}
The charge  is in the `adjoint' charge under the $usp(4)$ flavor symmetry,
and therefore $q_n=0$.
Therefore the 1-form symmetry is not extended.
